Discrete Quantum Theories and Computing

### 2018-03-17 dissertation.tex
dissertation.lyx "Chapter 3 Quantum Theories and Computing over Finite Fields" Updating...

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathComplex\mathbf{P}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{D}\mathComplex\mathbf{P}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}
\newcommand{\uf}{U_{\!f}}
\newcommand{\scalarPlus}{+}
\newcommand{\boolt}{\textsf{Bool}} 
\newcommand{\bfalse}{\texttt{\textbf{false}}}
\newcommand{\btrue}{\texttt{\textbf{true}}}
\newcommand{\dotprod}{dot product}
\newcommand{\Tr}{\ensuremath{\mathop{\mathrm{Tr}}\nolimits}}

@@

Assume the function $f$ is satisfiable at some input $a_{1},a_{2},\ldots,a_{n}$
denoted $\overline{a}$, and where $\ket{\overline{a}}=\ket{a_{1}}\ldots\ket{a_{n}}$.
In the second step, the state becomes $\ket{0}\ket{\overline{+}}$
as above. We can write this state as $\ket{0}\ket{\overline{a}}+\Sigma_{\overline{x}\neq\overline{a}}\ket{0}\ket{\overline{x}}$.
Applying $U_{f}$ produces $\ket{1}\ket{\overline{a}}+\Sigma_{\overline{x}\neq\overline{a}}\ket{0}\ket{\overline{x}}$.
We can rewrite this state as $\ket{+}\ket{\overline{a}}+\Sigma_{\overline{x}}\ket{0}\ket{\overline{x}}=\ket{+}\ket{\overline{a}}+\ket{0}\ket{\overline{+}}$,
where the summation is now over all vectors (notice that $\ket{0}\ket{\overline{a}}+\ket{0}\ket{\overline{a}}$
is the zero vector). Applying $S$ to each qubit in the second component
produces $\ket{+}\ket{\overline{S(a)}}+\ket{0}\ket{\overline{0}}$.
Applying $S^{\dagger}$ to the first component produces: $\ket{1}\ket{\overline{S(a)}}+\ket{0}\ket{\overline{0}}$.
Applying control-not gate, which applying $\sigma_{0}$ or $\sigma_{x}$
on the second component depending on the first component of the state,
and produces 
\begin{equation}
\ket{1}\left(\sigma_{x}\ket{\overline{S(a)}}\right)+\ket{0}\left(\sigma_{0}\ket{\overline{0}}\right)=\ket{1}\ket{\overline{\mathsf{not}\left(S(a)\right)}}+\ket{0}\ket{\overline{0}}\,.
\end{equation}
Applying $S^{\dagger}$ to the first component produces $\ket{+}\ket{\overline{\mathsf{not}\left(S(a)\right)}}+\ket{0}\ket{\overline{0}}$.
For the measurement of $\ket{+}\ket{\overline{\mathsf{not}\left(S(a)\right)}}+\ket{0}\ket{\overline{0}}$
to be guaranteed to never be $\ket{0}\ket{\overline{0}}$, we need
to verify that $\ket{+}\ket{\overline{\mathsf{not}\left(S(a)\right)}}$
has one occurrence $\ket{0}\ket{\overline{0}}$. This can be easily
proved as follows. Since each $a_{i}$ is either 0 or 1, then each
$S(a_{i})$ is either $+$ or $1$, and hence each $\mathsf{not}\left(S(a_{i})\right)$
is either~$+$ or~$0$. The result follows since any state with
a combination of $+$ and $0$, when expressed in the standard basis,
would consist of a superposition containing the state $\ket{0\ldots}$.

@@

The next task is to examine the consequences when we attempt to construct
$d$-dimensional vector spaces over the complexified fields $\ff{p^{2}}$
\cite{geom1}. (For readability, instead of writing column vectors,
we will often use the vector notation $\ket{\Psi}=(\alpha_{0}~\alpha_{1}~\ldots~\alpha_{d-1})^{T}$
and $\ket{\Phi}=(\beta_{0}~\beta_{1}~\ldots~\beta_{d-1})^{T}$, where
$(.)^{T}$ is the transpose of the \textit{row} vector $(.)$.) It
can be shown~\cite{HermitianProd} that, given two vectors 
\begin{equation}
\ket{\Psi}=\sum_{i=0}^{d-1}~\alpha_{i}\ket{i}\ ,\ \ket{\Phi}=\sum_{i=0}^{d-1}~\beta_{i}\ket{i},
\end{equation}
with scalars $\alpha_{i}$ and $\beta_{i}$ drawn from the field elements,
and orthonormal basis $\{\ket{i}\}$, the Hermitian \dotprod\ is
always reducible to the form \vspace{-0.1in}

\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{d-1}~\beta_{i}^{p}~\alpha_{i}^{\;}\ .\vspace{-0.1in}\label{innerprod}
\end{equation}

This product satisfies conditions A and B below, but not~C, because
in a finite field, addition can ``wrap around,'' making the concepts
of positive and negative meaningless and allowing the sum of non-zero
elements to be zero: 
\begin{itemize}
\item A. $\ip{\Phi}{\Psi}$ is the complex conjugate of $\ip{\Psi}{\Phi}$;
\vspace*{-0.25cm}

\item B. $\ip{\Phi}{\Psi}$ is conjugate linear in its first argument and
linear in its second argument; \vspace*{-0.25cm}

\item C. $\ip{\Psi}{\Psi}$ is always non-negative and is equal to~0 only
if $\ket{\Psi}$ is the zero vector. 
\end{itemize}
With just conditions A and B, it is possible to recover unitary operators,
and thus recover much of the relevant structure of Hilbert spaces
over the field of complex numbers. The failure of condition C, however,
plays havoc with the traditional notions of ordered probabilities
as well as the geometric notions of ordered distances and angles,
whose lengths and cosines, respectively, are normally expressed using
the inner product~\cite{geom}. In a separate development, we explore
the geometry of these finite fields and define a discrete version
of the Hopf fibration extending the Bloch sphere to $n$-qubits, as
well as determining discrete measures for the relative sizes of the
entangled, maximally entangled, and unentangled discrete states~\cite{geom1}.

\section{Vector Spaces over Complexified Finite Fields}

In order to address the intrinsic problems induced by the notion of
the continuum calculations of the previous section, one is led to
replace the infinite fields of CQC by discrete computable fields.
Accomplishing this while maintaining the essential elements of addition
and multiplication requires a brief excursion into the theory of fields,
and particularly the theory of finite fields.

Abstract algebra deals with various kinds of algebraic structures,
such as groups, rings, and fields, each defined by a different system
of axioms. A field~$\mathbf{F}$ is an algebraic structure consisting
of a set of elements equipped with the operations of addition, subtraction,
multiplication, and division \cite{numtheory.ref}. Fields may contain
an infinite or a finite number of elements. The rational $\mathbb{Q}$,
real $\mathbb{R}$, and complex numbers $\mathbb{C}$ are examples
of infinite fields, while the set $\mathbf{F}_{3}=\{0,1,2\}$ under
the usual multiplication and modular addition is an example of a finite
field. Finite fields are also known as Galois fields \cite{GT.ref}.

There are two distinguished elements in a field, the addition identity
$0$, and the multiplication identity $1$. Given the field $\mathbf{F}$,
the closed operations of addition, ``$+$'', and multiplication,
``$\gmult$'', satisfy the following set of axioms:
\begin{enumerate}
\item $\mathbf{F}$ is an Abelian group under the addition operation $+$
(additive group)
\item The multiplication operation $\gmult$ is associative and commutative.
The field has a multiplicative identity and the property that every
nonzero element has a multiplicative inverse
\item Distributive laws: For all $a,b,c\in\mathbf{F}$ 
\begin{eqnarray}
a\gmult(b+c)=a\gmult b+a\gmult c\\
(b+c)\gmult a=b\gmult a+c\gmult a\ .
\end{eqnarray}
\end{enumerate}
{}From now on, unless specified, we will omit the symbol $\gmult$
whenever we multiply two elements of a field.

Finite fields of $q$ elements, $\Fq=\{0,\ldots,q-1\}$, will play
a special role in this work. A simple explicit example is the following
addition and multiplication tables for $\mathbf{F}_{3}$: 
\[
\begin{array}{c|ccc}
+ & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 1 & 2\\
1 & 1 & 2 & 0\\
2 & 2 & 0 & 1
\end{array}\hspace*{2cm}\begin{array}{c|ccc}
\gmult & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 0 & 0\\
1 & 0 & 1 & 2\\
2 & 0 & 2 & 1
\end{array}
\]

Finite fields are classified by size. The characteristic of a field
is the least positive integer $m$ such that $m=1+1+1+\cdots+1\equiv0$,
and if no such $m$ exists we say that the field has characteristic
zero (which is the case for infinite fields such as $\mathbb{R}$).
It turns out that if the characteristic is non-zero it must be a prime
$p$. For every prime $p$ and positive integer $r$ there is a finite
field $\Fpx{p^{r}}$ of cardinality $q=p^{r}$ and characteristic
$p$ (Lagrange's theorem), which is unique up to field isomorphisms.
For every $a\in\Fq$, $a\neq0$, then $a^{q-1}=1$, implying the Frobenius
endomorphism (also a consequence of Fermat's little theorem) $a^{q}=a$,
which in turn permits us to write the multiplicative inverse of any
non-zero element in the field as $a^{-1}=a^{q-2}$, since $a^{q-2}a=a^{q-1}=1$.
Every subfield of the field $\Fq$, of cardinality $q=p^{r}$, has
$p^{r'}$ elements with some $r'$ dividing $r$, and for a given
$r'$ it is unique. Notice that a fundamental difference between finite
and infinite fields is one of topology: finite fields induce a compact
structure because of their modular arithmetic, permitting \textit{wrapping
around}, while that is not the case for fields of characteristic zero.
This feature may lead to fundamental physical consequences.

Consider the polynomial $x^{2}+1=0$ over a finite field $\Fp$. It
is known that this polynomial does not have solutions in the field
precisely when the prime $p$ is congruent to $3\imod{4}$ (see, e.g.,
\cite{numtheory.ref}).

For such primes, it is therefore possible to construct an extended
field $\Fpp$ whose elements are of the form $\alpha=a+ib$ with $a\in\Fp$,
$b\in\Fp$, and $i$ the root of the polynomial $x^{2}+1=0$. Since
the field elements $a+ib$ behave like discrete versions of the complex
numbers, we will refer to fields $\Fp$ with prime $p$ congruent
to $3\imod{4}$ as \emph{complexifiable finite fields}, and $i$-extended
fields $\Fpp$ with $p$ congruent to $3\imod{4}$ as \emph{complexified
finite fields}.

In a complexified finite field $\Fpp$, the Frobenius automorphism
that maps $\alpha\in\Fpp$ to $\alpha^{p}\in\Fpp$ acts like complex
conjugation. For example, in $\Fppx{3}$, we have $(2+i)^{3}=8+12i-6-i=2+11i$
which, in the field, is equal to $2-i$ since $11\equiv-1\imod{3}$.

@@

\begin{definition}{[}Hermitian dot product{]} Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in{\cal H}$, the Hermitian dot product of these vectors
is: 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}~\beta_{i}^{p}~\alpha_{i}^{\;}\ .
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in{\cal H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions A and B for inner products but violates condition C since
in every finite field there always exists a non-zero vector $\ket{\Psi}$
such that $\ip{\Psi}{\Psi}=0$. The reason is that addition in finite
fields eventually ``wraps around'' (because of their cyclic or modular
structure), allowing the sum of non-zero elements to be zero. The
fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$ decreases
with the order $p$.

@@

For vectors $\ket{\Psi}$ such that $\ip{\Psi}{\Psi}=\sum_{i=0}^{D-1}\alpha_{i}^{p}\,\alpha_{i}^{\;}=\sum_{i=0}^{D-1}\alpha_{i}^{p+1}=\sum_{i=0}^{D-1}|\alpha_{i}|^{2}$
has a square root in the field, one can define the following ``norm'':
\begin{equation}
||\Psi||=\sqrt{\ip{\Psi}{\Psi}}\ ,\label{norm-eq}
\end{equation}
which is valid only on a subspace of the field norm for finite fields.

%There is an alternative way of defining a proper inner product over a subset of the %field $\Fpp$. Consider prime numbers $p= 8 \prod_{l=1}^k q_l -1$, where  %$q_l$ are odd primes. This subset of the primes congruent to $3 \imod 4$ has the %special properti

\section{Irreducible Discrete $n$-qubit States: Generalized Discrete Bloch
Sphere}

\label{DQCnqubitBloch.sec}

In the one-qubit state with coefficients in $\Fpp$, the discrete
analog of the Bloch sphere is constructed by exact analogy to the
continuous case: we first require that the coefficients of the single
qubit basis obey 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1\label{DQA2.eq}
\end{equation}
in the discrete field. In \ref{appU.sec}, we show that there are
$p(p^{2}-1)$ such values. Given this requirement, which is similar
in form to the conservation of probability, but not as useful due
to the lack of orderable probability values, we can immediately conclude
that the discrete analog of the Hopf fibration is again 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\label{DQA6.eq}\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\ ,\nonumber 
\end{eqnarray}
but now with all computations in $(\mbox{mod}~p)$. At this point
one simply writes down all possible discrete values for the complex
numbers $\{\alpha_{0},\alpha_{1}\}$ satisfying Eq.~(\ref{DQA2.eq})
and enumerates those that project to the same value of $\{X,Y,Z\}$.
This equivalence class is the discrete analog of the circle in the
complex plane that was eliminated in the continuous case. In \ref{appH.sec},
we show that $p+1$ discrete values of $\{\alpha_{0},\alpha_{1}\}$
with unit norm map to the same point under the Hopf map \Eq{DQA6.eq};
we may think of these as discrete circles or projective lines of equivalent,
physically indistinguishable, complex phase. The surviving $p(p-1)$
values of $\{\alpha_{0},\alpha_{1}\}$ correspond to irreducible physical
states of the discrete single qubit system. %% Several examples are shown in Figure \ref{bloch.fig}. %%Thus,
for example, choosing the underlying field to be $\Fppx{3}$, there
are exactly 6 single-qubit state vectors to populate the Bloch sphere;
the four equivalent phase-multiples mapping to each of the six points
on the $\Fppx{3}$ Bloch sphere are collapsed and regarded as physically
indistinguishable. In Figure \ref{bloch.fig}, we plot the irreducible
states on the Bloch sphere for $p=3$, $7$, and $11$. Note that
the Cartesian lengths of the real vectors corresponding to the points
on the Bloch sphere vary considerably due to the nature of discrete
fields; we have artificially normalized them to a ``continuous world''
unit radius sphere for conceptual clarity.

\begin{figure}[htb]
%\vspace{-0.2in}  %% \includegraphics[width=0.9\columnwidth]{./Bloch_sphere.pdf}%%\includegraphics[width=2.0\columnwidth]{./picA.png}%%\includegraphics[width=2.0\columnwidth]{./picB.png}

\begin{centering}
\hspace*{-0.5cm}\includegraphics[width=1.1\columnwidth]{../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/picC.png}
%%\includegraphics[width=0.5\textwidth]{./picAB.eps}\vspace{-0.2in}
 %% adjust blank space?
\par\end{centering}
\caption{Schematically normalized plots of the elements of the discrete Bloch
sphere, the irreducible single-qubit (two-dimensional) state vectors
with unit norm over the field $\Fpp$. We show the results for $p=3$,
$7$, and $11$. For example, in $\Fppx{3}$, there are 24 vectors
of unit norm, but only the 6 inequivalent classes appear in the plot.
The $p+1=4$ equivalent vectors in each class differ only by a complex
discrete phase.}
\label{bloch.fig} 
\end{figure}

\subsection{Counting states on the $n$-qubit Bloch sphere}

We have the unique opportunity in the finite-field approach to quantum
computing to precisely identify and enumerate the physical states.
In the conventional theory, as we have seen, we employ a generalized
Hopf fibration on the normalized states to project out a circle of
phase-equivalent states, yielding the generalized Bloch sphere.

In the introduction to this section, we sketched the counting of the
irreducible single-qubit discrete states. To count the number of inequivalent
discrete states for the general $n$-qubit case with coefficients
in $\Fpp$, we first must find the set of unit-norm states, and then
determine the equivalence classes of unit-norm states under discrete
phase transformations; we can then enumerate the list of states on
the discrete generalized Bloch sphere. By executing computer searches
of these spaces, we discovered an hypothesis for a closed-form solution
for the counting of the states, and were then able to find a rigorous
inductive proof of the enumeration, which is presented in the Appendix.

This process of describing the discrete $n$-qubit irreducible states
can again be understood geometrically by following the discrete analog
of the Hopf fibration. First, we construct the discrete version of
the quadratic unit-length form that automatically annihilates the
distinction among states differing only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Hat{a}=(|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots)\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Hat{a}\cdot\Hat{a}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
{}From \ref{appH.sec}, we know that $p+1$ elements of this discrete
$\Sphere{2\times2^{n}-1}$ structure map to the \textit{same point\/}
in $\Hat{a}$. Each set of $(p+1)$ redundant points is, geometrically
speaking, the \textit{discrete Hopf fibration circle\/} living above
each \textit{irreducible\/} point of the $n$-qubit state description.
These $p+1$ points are interpretable as the $p$ finite points plus
the single point at infinity of the projective discrete line (see,
e.g., \cite{Arnold}).

The next part of this argument is the determination of the unit-norm
states, effectively the space of allowed discrete partitions of unity;
we cannot exactly call these ``probability-conserving'' sectors
of the state coefficients since we do not have a well defined notion
of probability, but we do have a well-defined notion of partition
of unity. The tally of unit-norm states is $p^{2^{n}-1}(p^{2^{n}}-1)$
(see \ref{appU.sec}) compared to the total number $p^{2\times2^{n}}$
of possible complex integer state vectors that could be chosen. This
unit-norm state structure is the discrete analog of $\Sphere{2\times2^{n}-1}$.

Finally, we repeat the last step of the $n$-qubit continuous Hopf
fibration process for discrete $n$-qubit states, eliminating the
discrete set of $p+1$ equivalent points that map to the same point
$\Hat{a}$ on the generalized $n$-qubit Bloch sphere. Dividing the
tally $p^{2^{n}-1}(p^{2^{n}}-1)$ of unit norm states by the $p+1$
elements of each phase-equivalent discrete circle, we find 
\[
\frac{p^{2^{n}-1}(p^{2^{n}}-1)}{p+1}=p^{2^{n}-1}(p-1)\,\prod_{k=1}^{n-1}(p^{2^{k}}+1)
\]
as the total count of unique irreducible states in a discrete $n$-qubit
configuration (see \ref{appI.sec}). The resulting object is precisely
the discrete version of $\CP{D-1}$, which we might call a \textit{discrete
complex projective space\/} or $\DCP{D-1}$, where $D=2^{n}$ as
usual.

\Comment{

\subsection{Counting states on the $n$-qubit Bloch sphere}

Although we do not have a rigorous proof for the counting of discrete
unit-norm states and the number of irreducible unit-norm states in
the discrete generalized Bloch sphere for $n$-qubit states with coefficients
in $\Fpp$, we have run extensive computer searches of the state features
for a range of compatible primes $p$ and number of qubits $n$. From
these experimental investigations, we have induced a closed-form solution
for the counting that agrees for every possible combination of parameters
that can be computed in reasonable time (the computational complexity
becomes intractable very quickly).

For $n$-qubit discrete states with coefficients in $\Fpp$, our calculation
begins with an enumeration of all possible $2^{n}$-tuples of state
coefficients in $\Fpp$; then the moduli-squared of these tuples are
summed and compared to unity in $(\mbox{mod}~p)$. Those that survive
are accumulated in a list of allowed discrete partitions of unity;
we cannot exactly call these ``probability-conserving'' sectors
of the state coefficients since we do not have a well defined notion
of probability, but we do have a well-defined notion of partition
of unity. The heuristic number that results for the tally of unit-norm
states is $p^{2^{n}-1}(p^{2^{n}}-1)$, compared to the total number
$p^{2\times2^{n}}$ of possible complex integer state vectors that
could be chosen. This unit-norm state structure is now the discrete
analog of $\Sphere{2\times2^{n}-1}$.

Finally, we construct the discrete version of the quadratic unit-length
form that automatically annihilates the distinction among states differing
only by a discrete phase, 
\begin{eqnarray}
\hspace*{-0.5cm}\Hat{a}=\{|\alpha_{i}|^{2},...,\sqrt{2}\ \mathrm{Re}\ \alpha_{i}\alpha_{j}^{*},\ldots,\sqrt{2}\ \mathrm{Im}\ \alpha_{i}\alpha_{j}^{*},\ldots\}\ ,
\end{eqnarray}
where 
\begin{eqnarray}
\Hat{a}\cdot\Hat{a}=\left(\sum_{i=0}^{D-1}\left|\alpha_{i}\right|^{2}\right)^{2}=1\ .
\end{eqnarray}
Computing all the elements of the discrete $\Sphere{2\times2^{n}-1}$
structure that map to the \textit{same point\/} in $\Hat{a}$, we
find a $(p+1)$-fold redundancy, exactly as for the one-qubit case;
each set of $(p+1)$ redundant points is, geometrically speaking,
the \textit{discrete Hopf fibration circle\/} (see, e.g., Arnold
\cite{Arnold}) living above each \textit{irreducible\/} point of
the $n$-qubit state description.

In summary, when the $n$-qubit continuous Hopf fibration process
is repeated for the discrete analog of $n$-qubit states, we find
that that analog of a phase ``circle'' is a discrete set $p+1$
equivalent points that map to the same point $\Hat{a}$ on the generalized
$n$-qubit Bloch sphere, and therefore there are a total of $p^{2^{n}-1}(p^{2^{n}}-1)/(p+1)$
unique irreducible states in a discrete $n$-qubit configuration.
The resulting object is precisely the discrete version of $\CP{D-1}$,
which we might call a \textit{discrete complex projective space\/}
or $\DCP{D-1}$. }

\section{Geometry of Entangled States}

Without regard to uniqueness, an $n$-qubit state with discrete complex
coefficients in $\Fpp$ will have the total possible space of coefficients
with dimension $p^{2\times2^{n}}$ (including the null state). Imposing
the condition of a length-one norm %% AMR: replaced 'inner-product' by 'norm'in
$\Fp$, this number is reduced to $p^{2^{n}-1}(p^{2^{n}}-1)$. The
ratio of all the states to the unit-norm states is asymptotically
$p$: 
\begin{eqnarray}
\frac{p^{2^{n}+1}}{p^{2^{n}}-1}\rightarrow p\ ,
\end{eqnarray}
so there are roughly $p$ sets of coefficients, for any number of
qubits $n$, that are discarded for each retained unit-length state
vector. A factor of $p+1$ more states are discarded in forming the
discrete Bloch sphere of irreducible states. Selected plots of the
full space compared to both the unit-norm space and the irreducible
space for a selection of complexified finite fields are shown in Figure
\ref{statePlot3Unit.fig} for 1, 2, 3, and 4 qubits. %Figure \ref{statePlot3Unit.fig} adds the size of the irreducible state%space at the bottom.

%% Source of this is%%  ~hansona/home2/papers/dqc-geom-2012/00mathFigs/A-Test-Galois-Nqubit.nb

%\begin{figure}[htb]%\begin{center}%\includegraphics[width=1.0\columnwidth]{./statePlot1.pdf}%%\vspace{-0.2in}  %% adjust blank space?%\end{center}% \caption{Logarithmic plot of the number of discrete unnormalized%states (top, in red), vs the number of normalized discrete states%(bottom, in blue) for the first 6 $\Fpp$-compatible primes,  (3, 7,%11, 19, 23, 31), for the number of qubits 1, 2, 3 and 4. }% \label{statePlotUnit.fig}%\end{figure}

\begin{figure}[htb]
\begin{centering}
\includegraphics[width=1\columnwidth]{../../../Public/Documents/StudyInIUB/Computer Science/201209-12.M800/Manuscripts/statePlot3.pdf}
%\vspace{-0.2in}  %% adjust blank space?
\par\end{centering}
\caption{Logarithmic plot of the number of discrete unnormalized states (top,
in red), vs the number of normalized discrete states (middle, in blue),
vs the irreducible states (bottom, in green) for the first 6 $\Fpp$-compatible
primes, (3, 7, 11, 19, 23, 31), for the number of qubits 1, 2, 3 and
4. }
\label{statePlot3Unit.fig} 
\end{figure}

\subsection{Unentangled vs Entangled Discrete States}

For a given $p$ and the corresponding complexified field $\Fpp$,
the $n$-qubit discrete quantum states with coefficients in $\Fpp$
can be classified by their degree of entanglement to a level of precision
that is unavailable in the continuous theory. We look first at the
unentangled $n$-qubit states, which are direct product states of
the form 
\begin{equation}
\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}\ .\label{dqcNqUntg.eq}
\end{equation}
Without regard to normalization, there are $(p^{4})^{n}$ possible
unentangled states out of the total of $p^{2\times2^{n}}$ states
noted above. When we normalize the individual product states to unit
norm, the norm of the entire $n$-qubit state becomes the product
of those unit norms, and is automatically normalized to one. We have
already seen that each single-qubit normalized state in the tensor
product \Eq{dqcNqUntg.eq} has precisely $p(p-1)$ irreducible components.

\subsection{Completely Unentangled States and the Discrete Bloch Sphere}

In effect, the irreducible states for unentangled $n$-qubit configurations
reduce to a single Bloch sphere for each one-qubit component $\ket{\psi_{j}}$,
and thus the whole set of states is defined by an $n$-tuple of discrete
Bloch sphere coordinates. Since each Bloch sphere in $\Fpp$ has $p(p-1)$
distinct irreducible components, we have 
\[
\mbox{{\bf Count of Unentangled States}}=p^{n}(p-1)^{n}\ .
\]

>>>>

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{}[section]

\newcommand{\muB}{\ensuremath{\mu^{B}}}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}\mathrel{}\middle|\mathrel{}{#2}\right\} }}
\newcommand{\gmult}{*}
\newcommand{\Fpx}[1]{\mathbb{F}_{{#1}}}
\newcommand{\Fp}{\Fpx{p}}
\newcommand{\Fppx}[1]{\Fpx{{#1}^2}}
\newcommand{\Fpp}{\Fppx{p}}
\newcommand{\Fq}{\Fpx{q}}
\newcommand{\ff}[1]{\Fpx{#1}}
\newcommand{\ffzx}[2]{\Fpx{#1}^{{#2}\;*}}
\newcommand{\ffx}[2]{\Fpx{#1}^{{#2}}}
\newcommand{\ffzd}[1]{\ffzx{#1}{d}}
\newcommand{\ffd}[1]{\ffx{#1}{d}}
\newcommand{\VVec}[1]{\vec{\bf #1}}
\newcommand{\mathReal}{\mathbb{R}}
\newcommand{\mathComplex}{\mathbb{C}}
\newcommand{\braket}[2]{\ip{#1}{#2}}
\newcommand{\todo}[1]{\textbf{TODO.~#1}}
\newcommand{\Eq}[1]{Eq.~(\ref{#1})}
\newcommand{\Sphere}[1]{\mathbf{S}^{#1}}
\newcommand{\R}{\mathReal}
\newcommand{\rme}{\mathrm{e}}
\newcommand{\rmi}{\mathrm{i}}
\newcommand{\CP}[1]{\mathComplex\mathbf{P}^{#1}}  % Complex projective space
\newcommand{\DCP}[1]{\mathbf{D}\mathComplex\mathbf{P}^{#1}} % Discrete complex projective
\def\fh{\mathfrak{h}}
\newcommand{\uf}{U_{\!f}}
\newcommand{\scalarPlus}{+}
\newcommand{\boolt}{\textsf{Bool}} 
\newcommand{\bfalse}{\texttt{\textbf{false}}}
\newcommand{\btrue}{\texttt{\textbf{true}}}
\newcommand{\dotprod}{dot product}
\newcommand{\Tr}{\ensuremath{\mathop{\mathrm{Tr}}\nolimits}}
\newcommand{\fnorm}[1]{\mathsf{N}(#1)}

@@

Assume the function $f$ is satisfiable at some input $a_{1},a_{2},\ldots,a_{n}$
denoted $\overline{a}$, and where $\ket{\overline{a}}=\ket{a_{1}}\ldots\ket{a_{n}}$.
In the second step, the state becomes $\ket{0}\ket{\overline{+}}$
as above. We can write this state as $\ket{0}\ket{\overline{a}}+\Sigma_{\overline{x}\neq\overline{a}}\ket{0}\ket{\overline{x}}$.
Applying $U_{f}$ produces $\ket{1}\ket{\overline{a}}+\Sigma_{\overline{x}\neq\overline{a}}\ket{0}\ket{\overline{x}}$.
We can rewrite this state as $\ket{+}\ket{\overline{a}}+\Sigma_{\overline{x}}\ket{0}\ket{\overline{x}}=\ket{+}\ket{\overline{a}}+\ket{0}\ket{\overline{+}}$,
where the summation is now over all vectors (notice that $\ket{0}\ket{\overline{a}}+\ket{0}\ket{\overline{a}}$
is the zero vector). Applying $S$ to each qubit in the second component
produces $\ket{+}\ket{\overline{S(a)}}+\ket{0}\ket{\overline{0}}$.
Applying $S^{\dagger}$ to the first component produces: $\ket{1}\ket{\overline{S(a)}}+\ket{0}\ket{\overline{0}}$.
Applying control-not gate, which applying $\sigma_{0}$ or $\sigma_{x}$
on the second component depending on the first component of the state,
and produces 
\begin{equation}
\ket{1}\left(\sigma_{x}\ket{\overline{S(a)}}\right)+\ket{0}\left(\sigma_{0}\ket{\overline{0}}\right)=\ket{1}\ket{\overline{\mathsf{not}\left(S(a)\right)}}+\ket{0}\ket{\overline{0}}\,.
\end{equation}
Applying $S^{\dagger}$ to the first component produces $\ket{+}\ket{\overline{\mathsf{not}\left(S(a)\right)}}+\ket{0}\ket{\overline{0}}$.
For the measurement of $\ket{+}\ket{\overline{\mathsf{not}\left(S(a)\right)}}+\ket{0}\ket{\overline{0}}$
to be guaranteed to never be $\ket{0}\ket{\overline{0}}$, we need
to verify that $\ket{+}\ket{\overline{\mathsf{not}\left(S(a)\right)}}$
has one occurrence $\ket{0}\ket{\overline{0}}$. \todo{The following
need to be rewritten.}This can be easily proved as follows. Since
each $a_{i}$ is either 0 or 1, then each $S(a_{i})$ is either $+$
or $1$, and hence each $\mathsf{not}\left(S(a_{i})\right)$ is either~$+$
or~$0$. The result follows since any state with a combination of
$+$ and $0$, when expressed in the standard basis, would consist
of a superposition containing the state $\ket{0\ldots}$.

@@

In a complexified finite field $\Fpp$, the Frobenius automorphism
that maps $\alpha\in\Fpp$ to $\alpha^{p}\in\Fpp$ acts like complex
conjugation. For example, in $\Fppx{3}$, we have $(2+i)^{3}=8+12i-6-i=2+11i$
which, in the field, is equal to $2-i$ since $11\equiv-1\pmod{3}$.

@@

\begin{definition}[Hermitian dot product] Given vectors $\ket{\Phi}$
and $\ket{\Psi}\in{\cal H}$, the Hermitian dot product of these vectors
is: 
\begin{equation}
\ip{\Phi}{\Psi}=\sum_{i=0}^{D-1}~\beta_{i}^{p}~\alpha_{i}^{\;}\ .
\end{equation}
\end{definition} Two vectors $\ket{\Phi}$ and $\ket{\Psi}\in{\cal H}$
are said to be orthogonal if $\ip{\Phi}{\Psi}=0$. This product satisfies
conditions A and B for inner products but violates condition C since
in every finite field there always exists a non-zero vector $\ket{\Psi}$
such that $\ip{\Psi}{\Psi}=0$. The reason is that addition in finite
fields eventually ``wraps around'' (because of their cyclic or modular
structure), allowing the sum of non-zero elements to be zero. The
fraction of non-zero vectors satisfying $\ip{\Psi}{\Psi}=0$ decreases
with the order $p$.

@@
